### 2019-08-16 boek.tex
update document
update manuscript
update texs and files
use zotero instead
fix logos
update
update figs, texs, and other files

\documentclass[11pt,a4paper]{book}

@@

% adjust margins (should be done before importing fancyhdr!)
\usepackage[margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm

% text packages
\usepackage[british]{babel}
\usepackage{setspace}

@@

% style packages
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

@@

% figure packages
\usepackage{tikz,pgfplots}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\graphicspath{{fig/}}
\usepackage{xcolor}

@@

% table packages
\usepackage{tabularx}
\usepackage{tabulary}

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}

@@

@@

@@

% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}
\hypersetup{%
	pdftitle = {A title},
	pdfsubject = {A subject},
	pdfkeywords = {import keywords},
	pdfauthor = {Me of course},
	pdfcreator = {\LaTeX},
}

% Set style variables
\setlength{\parindent}{0cm} 
\renewcommand{\contentsname}{Table of Contents}
\renewcommand{\listfigurename}{List of Figures}
\renewcommand{\listtablename}{List of Tables}
\renewcommand{\appendixname}{}

@@

%\onehalfspace

@@

% geen paginanummering tot we aan de inhoudsopgave komen
\pagenumbering{alph}

@@

% tabel van afkortingen en symbolen

% hoofdstukken
\mainmatter
\pagestyle{mainmatter}

\include{ch1_singlecellbiology}

@@

@@

\include{ch5_incgraph}

\include{ch6_brood}

@@

% appendices

\appendix

\printbibliography

\backmatter

% eventueel: lijst van figuren en tabellen
%\listoffigures
%\listoftables

% lege pagina (!!)

>>>>

\documentclass[11pt,hidelinks]{book}

@@

% Use Unicode characters when possible
%\usepackage{xltxtra}

%\usepackage[default,scale=0.95]{opensans}
%\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}

% adjust margins (should be done before importing fancyhdr!)
\usepackage[paperheight=24cm,paperwidth=16cm,margin=2.5cm,a4paper]{geometry} %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm

% text packages
\usepackage{csquotes}
\usepackage[english]{babel}
\usepackage{setspace}

@@

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
%\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

@@

% figure packages
\usepackage{tikz}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\usepackage{xcolor}
\usepackage{colortbl}

@@

% table packages
\usepackage{tabularx}
\usepackage{tabulary}
\usepackage{booktabs}

% sections in allcaps
\usepackage{titlesec}
%\titleformat{\section} {\Large\bfseries\center}{}{0em}{\hrule height 0.4pt\vspace{2pt}}[\hrule height 0.4pt] \titlespacing*{\section} {0pt}{0pt}{10pt}
%\titleformat{\subsection} {\large\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsection} {0pt}{0pt}{0pt}
%\titleformat{\subsubsection} {\normalsize\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsubsection}  {0pt}{0pt}{0pt}
%\titleformat{\section} [block]{\Large\bfseries\filcenter}{}{0em}{\MakeUppercase}
%\titleformat{\subsection} [block]{\large\bfseries\filcenter}{}{0em}{\MakeUppercase}
%\titleformat{\subsubsection} [block]{\normalsize\bfseries\filcenter}{}{0em}{\MakeUppercase}

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}
%
%% substitute in references
%\DeclareDatafieldSet{setall}{
%	\member[datatype=literal]
%	\member[datatype=name]
%	\member[field=journal]% journal is special since it is
%	% actually journaltitle
%}
%
%\DeclareSourcemap{
%	\maps[datatype=bibtex]{
%		\map[overwrite, foreach={setall}]{
%			% \`{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0300}},
%			replace=\regexp{\x{00EC}}]
%			% \'{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0301}},
%			replace=\regexp{\x{00ED}}]
%			% \^{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0302}},
%			replace=\regexp{\x{00EE}}]
%			% \"{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0308}},
%			replace=\regexp{\x{00EF}}]
%		}
%	}
%}

@@

% add checkmark commands
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

% fix named refs for starred sections
\makeatletter
\def\ttl@useclass#1#2{%
	\@ifstar
	{\ttl@labeltrue\@dblarg{#1{#2}}}% {\ttl@labelfalse#1{#2}[]}%
	{\ttl@labeltrue\@dblarg{#1{#2}}}}
\makeatother

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE FIGURE HELPERS %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\hugefigure}{\textwidth}
\newcommand{\LARGEfigure}{0.8\textwidth} % 1 per col
\newcommand{\LARgefigure}{0.6\textwidth} % 1 per col
\newcommand{\Largefigure}{0.5\textwidth} % 1 per col
\newcommand{\largefigure}{0.37\textwidth} % 2 per col
\newcommand{\normalfigure}{0.3\textwidth} % 3 per col
\newcommand{\smallfigure}{0.24\textwidth} % 3 per col
\newcommand{\Smallfigure}{0.19\textwidth} % 4 per col
\newcommand{\SMALLfigure}{0.15\textwidth} % 5 per col

\makeatletter
\newcommand*{\centerfloat}{%
	\parindent \z@
	\leftskip \z@ \@plus 1fil \@minus \textwidth
	\rightskip\leftskip
	\parfillskip \z@skip}
\makeatother

\newcommand{\picfig}[4]{
\begin{figure}[ht!]
	\centering \centerfloat
	\includegraphics[width=#2]{#1}
	\caption{#4}
	\label{#3}
\end{figure}}

%\newcommand{\picsubf}[2]{
%	\begin{subfigure}[b]{#2}
%		\fbox{\includegraphics[width=\linewidth]{#1}}
%\end{subfigure}}
%
%\newcommand{\picsubfc}[4]{
%	\begin{subfigure}[b]{#4}
%		\fbox{\includegraphics[width=\linewidth]{#1}}
%		\caption{#2}
%		\label{#3}
%\end{subfigure}}
%
%\newcommand{\picsub}[2]{
%	\begin{subfigure}[b]{#2}
%		\includegraphics[width=\linewidth]{#1}
%\end{subfigure}}
%
%\newcommand{\picsubc}[4]{
%	\begin{subfigure}[b]{#4}
%		\includegraphics[width=\linewidth]{#1}
%		\caption{#2}
%		\label{#3}
%\end{subfigure}}

@@

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}

% Set style variables
%\setlength{\parindent}{0cm} 
\renewcommand{\contentsname}{Table of Contents}
\renewcommand{\listfigurename}{List of Figures}
\renewcommand{\listtablename}{List of Tables}
\renewcommand{\appendixname}{}
\renewcommand{\baselinestretch}{1.25}

\usepackage{fontspec}

\setmainfont [Path = fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

@@

@@

\pagenumbering{alph}

@@

% chapters
\mainmatter
\pagestyle{mainmatter}

\include{ch1_introduction}

% landmark mds
% princurve
% GillespieSSA
% dyngen

@@

% include gng ?

@@

\include{ch5_brood}

\include{ch6_incgraph}

@@

\include{chX_backmatter}

### 2019-08-16 ch1_introduction.tex
update figs, texs, and other files

Adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
{\footnotesize $^*$ Equal contribution}
\newpage

@@

Transformative technological advances have enabled the profiling of single cells in a high-throughput approach, thus creating room for single-cell omics (Figure \ref{fig:profiling_convergence}). The first such technologies allowed to profile the whole transcriptome of thousands of single cells. Current developments in the field allow for much greater cell counts, to also measure proteomic expression levels, or even employ imaging techniques to also capture the spatial information of each molecule within a cell.

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/introduction/profiling_convergence} % TODO: change image
	\caption{
		\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is ex-tending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. % TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the Human Cell Atlas. The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells. 
\section{Machine learning}
The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles, which is something that biologists have relied upon for decades. Instead, biologists have to rely on computer scientists to develop new software to be able to extract meaningful information from that data. Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.

\subsection{Dimensionality reduction}

\subsection{Trajectory inference}

@@

\section{Research objectives}

>>>>

Partially adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
{\footnotesize $^*$ Equal contribution}
\newpage

@@

Transformative technological advances have enabled the profiling of single cells in a high-throughput approach, thus creating room for single-cell omics (Figure \ref{fig:profiling_convergence}). The first such technologies allowed to profile the whole transcriptome of thousands of single cells. Current developments in the field allow for much greater cell counts, to also measure other levels of information such as proteomic expression and spatial information. 
% maybe use this image? https://github.com/arnavm/multimodal-scRNA-seq

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/introduction/profiling_convergence} % TODO: change image
	\caption{
		\textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. % TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the Human Cell Atlas. The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.

\section{Computational tools}
The sheer size of the datasets generated by single-cell omics technologies disallows for manual analysis of the individual profiles. However, it opens up novel opportunities for collaborations between biologists and computer scientists to develop new computational tools to infer models of dynamic processes in an automated and unbiased fashion. This section introduces various types of computational tools which are of particular interest when analysing single-cell omics data \footnote{Yet their applications and implications are much more comprehensive than only single-cell omics.}.
\subsection{Dimensionality reduction}
The dimensionality of single-cell omics datasets is typically much too high for most modelling algorithms to tackle directly. Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful representation with fewer dimensions. DR methods construct this representation by reducing the gene space, the cell space, or both while attempting to maintain the main characteristics of the original data.
% Could use a figure here.
Ideally, the reduced representation should have a dimensionality that corresponds to the intrinsic dimensionality of the biological system.

There are many ways of classifying DR methods \cite{engel_surveydimensionreduction_2012}, but this work will use the following main categories: feature projection-based and manifold learning. 
Projection-based DR methods aim to perform a linear transformation of the data while preserving the pairwise distances between samples as much as possible. Examples of commonly used projection-based DR methods in single-cell omics are PCA, MDS, LMDS. 
Manifold learning methods are methods which reconstruct a higher-order structure in the original space (e.g. a graph or a grid), visualising the structure in a lower-dimensional space, and mapping the original samples to the lower-dimensional space. Manifold learning can be an iterative optimisation process using a predefined criterion. Examples of manifold learning techniques are t-SNE, Diffusion Maps and UMAP.

% For more information, we refer the reader to other reviews for a thorough discussion of the advantages and disadvantages of different dimensionality reduction methods [37, 38].
\subsection{Trajectory inference}
Single-cell omics data provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modelled computationally using trajectory inference (TI) methods which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}.

The initial dataset can be either a single snapshot of a mixture of cells in different stages or a set of samples collected at different time points (Figure \ref{fig:trajectory_inference}A). 
TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process \cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017} (Figure \ref{fig:trajectory_inference}B). Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

\begin{figure}
	\centering\includegraphics[width=\Largefigure]{fig/introduction/trajectory_inference_slim} % TODO: Update figure to match text!!!!
	\caption{
		Applications of single-cell trajectory inference methods. (A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations. (B) UPDATE!
	}
	\label{fig:trajectory_inference}
\end{figure}

A plethora of TI methods have been developed over the past few years, and still, more are created every month (Supplementary Table 1). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the "awesome-single-cell" list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. Each method has its own unique set of characteristics in terms of the underlying algorithm, required prior information and produced outputs. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. Therefore, these methods mainly focus on correctly ordering the cells along transitions in a fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

@@

\section{Research objectives}
Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
